Classless Java

### 2015-11-10 paper/sections/Compare.tex
related work (in progress)
Some work on comparision and related work. Section Future work now exists

\section{Comparing to traits and mixins}

\begin{comment}
Haoyuan

@@

Our approach is based on code generation by Java annotations. The model we generate encourages composability and reusability in object-oriented programming, and is considered to be an alternative to traits or mixins,  meanwhile achieving better performance in some situations. Hence it is necessary for us to make a comparison between this approach and traits (or mixins) we commonly used before.

Our approach is quite different from mixins, in the sense that we use the trait model of explicitly resolving conflicts. Just as [Scharli2003] demonstrated in the paper, mixin inheritance is a good approach of achieving code reuse, nevertheless, the mixin model is not so expressive to resolve conflicts from many mixins. In the trait model, aliases and exclusions are provided for explicit conflict resolution. Such operations can actually be modelled in the mechanism of our approach. On the other hand, the total ordering property [Scharli2003] restricts the composition of units of behavior. The linearization of mixin inheritance cannot provide a perfect total ordering for various requirements, however, in Java, we have multiple inheritance of interfaces, where methods are inherited at the same time from super interfaces.

Here we present how the original operations on traits are supported by our model.
\begin{itemize}
\item \textbf{Symmetric sum}: the symmetric composition of two disjoint traits is achieved by simply implementing two interfaces in Java correspondingly, without overriding any method. The composition relies on multiple inheritance on interfaces, which is supported by Java. Below is a simple example.
    \begin{lstlisting}
    interface A { int x(); }
    interface B { int y(); }
    interface C extends A, B {}
    \end{lstlisting}
\item \textbf{Override}: the overriding operation (also known as asymmetric sum) is modelled by implementing many interfaces, while overriding some methods inside. The code below gives an example of explicitly specifying which super interface to refer to, regarding two methods with the same name.
    \begin{lstlisting}
    interface A { default int m() {return 1;} }
    interface B { default int m() {return 2;} }
    interface C extends A, B { default int m() {return B.super.m();} }
    \end{lstlisting}
    Here the method \texttt{m()} in interface \texttt{C} simply inherits from \texttt{B.m()}.
\item \textbf{Alias}: an alias operation adds a new name to an old method when creating the new trait. In Java, we just create a new method with reference to the existing method in its super interface. See the example below, where the new method \texttt{k()} is an alias of the existing method \texttt{m()}.
    \begin{lstlisting}
    interface A { default int m() {return 1;} }
    interface B extends A { default int k() {return A.super.m();} }
    \end{lstlisting}
\item \textbf{Exclusion}: exclusion is also supported in Java, where method declarations can hide the default methods correspondingly in the super interfaces. See the example below.
    \begin{lstlisting}
    interface A { default int m() {return 1;} }
    interface B extends A { int m(); }
    \end{lstlisting}
\end{itemize}

Besides, we support more features than the original trait model:
\begin{itemize}
\item We provide \texttt{of} methods for the annotated interfaces. During annotation processing time, the ``fields'' inside an interface are observed and a static method \texttt{of} is automatically injected to the interface with its arguments correspondingly. Such a method is a replacement to the constructors in original traits, making instantiation more convenient to use.
\item We provide \texttt{with-} and \texttt{clone} methods as auxiliary constructors. A \texttt{with-} method is generated for each field, just like record update, returning the new object with that field updated. A \texttt{clone} method is generated for the interface, returning a copy of the current object. Furthermore, we do automatic return type refinement for these two kind of methods. This feature is comparatively useful in big examples, making operations and behaviors more flexible, which we will demonstrate later.
\item We provide two options for generating setters. There are two kind of setters which are commonly used, namely \textit{void setters} and \textit{fluent setters}. The only difference is that a fluent setter returns the object itself after setting, thus supporting a pipeline of such operations. The generation depends on which type of setter is declared in the interface by users.
\end{itemize}

@@

>>>>

%\section{Comparing to traits and mixins}\label{sec:comparison}
\subsection{Multiple Inheritance in ClassLess Java}
\begin{comment}
Haoyuan

@@

Our approach is based on forbidding the use of certain language constructs (classes), in order to rely on the modular composition offered by the rest of the language (interfaces).
Since Java was designed for classes, a direct class-less programming style is verbose and feel unnatural. However, annotation driven code generation is enough to overcome this difficulty, and
the resulting programming style encourages modularity, composability and reusability, by keeping a strong object oriented feel.
We consider our approach as an alternative to traits or mixins.

We now compare our approach to the original operations on traits, showing how are supported by our model.\footnote{This comparison shares  similarities with the encoding shown in \cite{bono14};
we differ in the 
proposed representation of
 \textbf{exclusion}.
The author of ~\cite{bono14} agree that our revised version is cleaner, typesafe and more direct.
}

\begin{itemize}
\item \textbf{Symmetric sum}: the symmetric composition of two disjoint traits is achieved by simply implementing two interfaces in Java correspondingly, without overriding any method. The composition relies on multiple inheritance on interfaces, which is supported by Java. Below is a simple example.
    \begin{lstlisting}
    interface A { int x(); }
    interface B { int y(); }
    interface C extends A, B {}
    \end{lstlisting}
\item \textbf{Override}: the overriding operation (also known as asymmetric/preferential sum) is modelled by implementing many interfaces, while overriding some methods inside. The code below gives an example of explicitly specifying which super interface take precedence, regarding conflict on a specific method.
    \begin{lstlisting}
    interface A { default int m() {return 1;} }
    interface B { default int m() {return 2;} }
    interface C extends A, B { default int m() {return B.super.m();} }
    \end{lstlisting}
    Here the method \texttt{m()} in interface \texttt{C} simply inherits from \texttt{B.m()}.
\item \textbf{Alias}: an alias operation adds a new name to an old method when creating the new trait. In Java, we just create a new method with reference to the existing method in its super interface. See the example below, where the new method \texttt{k()} is an alias of the existing method \texttt{m()}.
    \begin{lstlisting}
    interface A { default int m() {return 1;} }
    interface B extends A { default int k() {return A.super.m();} }
    \end{lstlisting}
\item \textbf{Exclusion}: exclusion is also supported in Java, where method declarations can hide the default methods correspondingly in the super interfaces. See the example below.
    \begin{lstlisting}
    interface A { default int m() {return 1;} }
    interface B extends A { int m(); }
    \end{lstlisting}
\end{itemize}

Besides, we support more features than the original trait model:
\begin{itemize}
\item We provide \texttt{of} methods for the annotated interfaces. During annotation processing time, the ``fields'' inside an interface are observed and a static method \texttt{of} is automatically injected to the interface with its arguments correspondingly. Such a method is a replacement to the constructors in original traits, making instantiation more convenient to use.
That is, in our approach there are only interfaces, our model requires a single concept, while the trait model requires traits \emph{and} classes.

\item We provide \texttt{with-} methods as auxiliary constructors. A \texttt{with-} method is generated for each field, just like record update, returning the new object with that field updated
%. A \texttt{clone} method is generated for the interface, returning a copy of the current object.
Furthermore, we do automatic return type refinement for these kind of methods. This feature is comparatively useful in big examples, making operations and behaviours more flexible.%, which we will demonstrate later.
\item We provide two options for generating setters. There are two kind of setters which are commonly used, namely \textit{void setters} and \textit{fluent setters}. The only difference is that a fluent setter returns the object itself after setting, thus supporting a pipeline of such operations. The generation depends on which type of setter is declared in the interface by users.
\end{itemize}

@@

There are other limitations of our current approach, but they may be addressed in future work. See \marco{reference to future work when there}

### 2015-11-10 paper/sections/Overview.tex
Some work on comparision and related work. Section Future work now exists

Just to give you a feeling on how much boring code \mixin is generating,  we show the correspondent code without then \mixin annotation.\footnote{Can you spot our intentional typo? This kind of code is so boring that our brain refuse to process it. Programmers should not be required to write this kind of code.}
\begin{lstlisting}
interface Point3D extends Point2D{
  Point3D withX(int val);
  Point3D withY(int val);
  Point3D withZ(int val);
  Point3D with(Point2D val);
  public static Point3D of(int x, int y, int z){return new Point3D(){
    public int x(){return x;} 
    public int y(){return y;}
    public int z(){return z;}
    public Point3D withX(int val){return Point3D.of(val,this.y(),this.z());}
    public Point3D withY(int val){return Point3D.of(this.x(),val,this.z());}
    public Point3D withZ(int val){return Point3D.of(this.x(),this.y(),val);}
    public Point3D with(Point2D val){
      if(val instanceof Point3D){return (Point3D)val;}
      return Point3D.of(val.x(),val.y(),this.z());
      }
    };
  }
}
\end{lstlisting}
Note how \Q@with-@ methods are automatically refined in their return type, so that code like 
\Q@Point3D p=Point3D.of(1,2,3); p=p.withX(42);@ will be accepted and behave as expected. If the programmer wishes to suppress this behaviour and keep the signature as it was, it is sufficient to redefine the \Q@with-@ methods in the new class repeating the old signature.
Again, the philosophy is that if the programmer provides something directly, \mixin do not touch it.
\Q@with-@ methods (functionally) update a field/property at a time.
This can be inefficient, and sometime hard to maintain.
Often we want to update many fields at the same time, for example using another object as source.
Following this idea, the method \Q@with(Point2D)@ is an example of a (functional) properties updater: 
it takes a certain type and in the current object update all field in that type that match fields in the current type.
The idea is that we want as result something that is still like \Q@this@, but modified to be as much as possible similar to the parameter.
The cast in \Q@with(Point2D)@ is trivially safe since it is guarded by an \Q@instanceof@ test. The idea is that if the parameter is a subtype of the current exact type, then we can just return the parameter, as something that is just ``more'' than \Q@this@.

@@

\lstinputlisting[linerange=80-92]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=BIRD

>>>>

Just to give you a feeling on how much boring code \mixin is generating,  we show the correspondent code without then \mixin annotation\footnote{
As you can see, the generated code is very repetitive, writing such code by hand can easily induce bugs, for example a distract programmer may
swap the arguments of one of the many calls of \Q@Point3D.of@.
}.\begin{lstlisting}
interface Point3D extends Point2D{
  Point3D withX(int val);
  Point3D withY(int val);
  Point3D withZ(int val);
  Point3D with(Point2D val);
  public static Point3D of(int x, int y, int z){return new Point3D(){
    public int x(){return x;} 
    public int y(){return y;}
    public int z(){return z;}
    public Point3D withX(int val){return Point3D.of(val,this.y(),this.z());}
    public Point3D withY(int val){return Point3D.of(this.x(),val,this.z());}
    public Point3D withZ(int val){return Point3D.of(this.x(),this.y(),val);}
    public Point3D with(Point2D val){
      if(val instanceof Point3D){return (Point3D)val;}
      return Point3D.of(val.x(),val.y(),this.z());
      }
    };
  }
}
\end{lstlisting}
Note how \Q@with-@ methods are automatically refined in their return type, so that code like 
\Q@Point3D p=Point3D.of(1,2,3); p=p.withX(42);@ will be accepted and behave as expected. If the programmer wishes to suppress this behaviour and keep the signature as it was, it is sufficient to redefine the \Q@with-@ methods in the new class repeating the old signature.
Again, the philosophy is that if the programmer provides something directly, \mixin do not touch it.
\Q@with-@ methods (functionally) update a field/property at a time.
This can be inefficient, and sometime hard to maintain.
Often we want to update many fields at the same time, for example using another object as source.
Following this idea, the method \Q@with(Point2D)@ is an example of a (functional) properties updater: 
it takes a certain type and in the current object update all field in that type that match fields in the current type.
The idea is that we want as result something that is still like \Q@this@, but modified to be as much as possible similar to the parameter.
The cast in \Q@with(Point2D)@ is trivially safe since it is guarded by an \Q@instanceof@ test. The idea is that if the parameter is a subtype of the current exact type, then we can just return the parameter, as something that is just ``more'' than \Q@this@.

@@

### 2015-11-10 paper/sections/RelatedWork.tex
Some work on comparision and related work. Section Future work now exists

\subsection{Multiple Inheritance in Java}
Since Java 8 default methods are introduced, which enables concrete method
implementation to be defined (via the \textbf{default} keyword) inside
interface. Since in Java supports extension on multiple interfaces (instead of
classes), the introduction of default methods opens the gate of doing multiple
inheritance in Java by using interfaces. Details on achieving multiple
inheritance by mimicking trait programing through interfaces are summarized by
Bono.et. al.~\cite{bono14}. There is also formalization proposed by
Goetz and Field~\cite{goetz12fdefenders} to formalize defender (default) methods
in Java. However this formalization is limited since they only model exactly one
method inside classes/interfaces, where in our formalization multiple methods
may be defined.

\yanlin{need to discuss: whether the comments for FTJ is fair.}
There are proposals for extending Java with traits. For example, FeatherTrait
Java (FTJ)~\cite{Liquori08ftj} by Liquori et al. extends the calculus of
Featherweight Java (FJ)~\cite{Igarashi01FJ} with statically-typed traits, adding
trait-based inheritance in Java. However, language extensions (including FTJ)
have a natural drawback: the programmer has to learn new syntax. In contrast,
our approach is completely compatible with the current Java language, so that
programmers don't need to pay any learning cost to adapt to this new classless
programming pattern. Another drawback which is particular for FTJ is that FTJ
doesn't have type for traits, hence the correctness check of trait is done when
type-checking classes. This choice makes the design of FTJ simpler but lost
typechecking efficiency (one trait will be potentially checked multiple times if
it is used in multiple classes).

\subsection{Multiple Inheritance in General}
Multiple inheritance is a great feature to have in a programming language, but
it is difficult to model and implement. These difficulties include the famous
diamond (fork-join) problem, conflicting methods, yo-yo problem, etc. In order
to model multiple inheritance, lots of models have been proposed in the past few
years, including mixins~\cite{bracha90mixin} and traits~\cite{scharli03traits}.
Both traits and mixins provide novel models of new programming architecture in
object oriented programming paradigm.

1) Mixins model named components that can be applied to various classes as
reusable functionality units. However, the well-known limitation of mixins is
that they suffer from linearization constraints, which hinders the
expressiveness of mixins and the ability of resolving conflicts.  As shown in
section~\ref{sec:comparison}, although they solved some multiple inheritance
problems, they are extremely restricting expressiveness.

2) Traits are a novel model to differentiate two functionalities out of classes:
units of reusable components and object factories. In trait models, traits are
responsible for units of reusable components, which contain pure methods as
reusable functionalities; whereas classes act as object factories, which
requires various functionalities from multiple traits. However drawbacks of
traits are obvious: traits do not allow state; traits can't be instantiated
directly.

3) C++ and Scala also try to provide solutions to multiple inheritance, but both
suffer from object initialization problems. Virtual inheritance in C++ provides
another solution to multiple inheritance (especially the diamond problem by
keeping only one copy of the base class)~\cite{ellis1990annotated}, however
suffers from object initialization problem as pointed out by Malayeri et
al.~\cite{malayeri2009cz}. It bypasses all constructor calls to virtual
superclasses, which would potentially cause serious semantic error. Scala
solution (traits) avoids this problem by disallowing constructor parameters,
causing no ambiguity in cases such as diamond problem. The drawback of this
elegant design is limited expressiveness.

@@

>>>>

\subsection{Formalization of Java8}
\marco{this go later, when we compare our formalization}
There is also formalization proposed by
Goetz and Field~\cite{goetz12fdefenders} to formalize defender (default) methods
in Java. However this formalization is limited since they only model exactly one
method inside classes/interfaces, where in our formalization multiple methods
may be defined.

%\yanlin{need to discuss: whether the comments for FTJ is fair.}

%No, was not fair, I know that work, it does modular type checking, and 
%more work after that they extend with more safe/modular typing things.
%also, Java8 IS a language extension on Java
%However, language extensions (including FTJ)
%have a natural drawback: the programmer has to learn new syntax. In contrast,
%our approach is completely compatible with the current Java language, so that
%programmers don't need to pay any learning cost to adapt to this new classless
%programming pattern. Another drawback which is particular for FTJ is that FTJ
%doesn't have type for traits, hence the correctness check of trait is done when
%type-checking classes. This choice makes the design of FTJ simpler but lost
%typechecking efficiency (one trait will be potentially checked multiple times if
%it is used in multiple classes).

\subsection{Multiple Inheritance in Object oriented languages}
Many authors have argued in favour or against multiple inheritance.
It provides expressive power, but it is difficult to model, and implement and can create programs that are hard to reason about.
These difficulties include the famous
diamond (fork-join) problem, conflicting methods, and the yo-yo problem.
\marco{add citation on the text over}

To conciliate the need for expressive power and the need for simplicity, 
lots of models have been proposed in the past few
years, including mixins~\cite{bracha90mixin} and traits~\cite{scharli03traits}.
They provide novel programming architecture models in the
OO paradigm. 
\begin{enumerate}
\item  Mixins allows to name components that can be applied to various classes as
reusable functionality units. However,  they suffer from linearisation: the order of mixin application is relevant in often subtitle and undesired ways. constraints. This hinders their
usability and their ability of resolving conflicts:
 the linearisation (total ordering) of mixin inheritance cannot provide a satisfactory resolution in all cases and restricts the flexibility of mixin composition.

 To fight those limitations, an algebra of mixin operators is introduced~\cite{CMS elena zucca davide ancona}, but this raised the complexity of the approach, especially when constructors and fields are considered~\cite{Ecoop FJigsaw marco}.

\item 
Simplifying the mixin algebras approach, traits draw a strong line between units of reuse (traits) and object factories (classes)
In this model, traits~\cite{Scharli2003} units of reusable code,  containing only methods as
reusable functionalities. Thus, no state/state initialization is considered.

Classes act as object factories, requiring functionalities from multiple traits.
Traits offers a trait algebra  with operations like sum, alias and exclusion, provided for explicit conflict resolution.

Concluding, (pure) traits do not allow state and they do not offers any reuse instrument to ensure that state is coherently initialized when finally defined in classes.
Traits can't be instantiated. This model requires two concepts (traits and classes) to coexist and cooperate.

Some authors see this as good language design fostering good software development by helping programmers to think about the structure of their programs.
However, other authors see the need of two concepts and the absence of state as drawbacks of this model.

\item C++ and Scala also try to provide solutions to multiple inheritance, but both
suffer from object initialization problems. Virtual inheritance in C++ provides
another solution to multiple inheritance (especially the diamond problem by
keeping only one copy of the base class)~\cite{ellis1990annotated}, however
suffers from object initialization problem as pointed out by Malayeri et
al.~\cite{malayeri2009cz}. It bypasses all constructor calls to virtual
superclasses, which would potentially cause serious semantic error. Scala
solution (very similar to linearised mixins, but misleadingly called traits in the language) avoids this problem by disallowing constructor parameters,
causing no ambiguity in cases such as diamond problem. 
This approach has limited expressiveness, and suffers from all the problems of linearised mixin composition.
Python also offers multiple inheritance via linearised mixins. Indeed python any class is implicitly a mixin, and mixin composition informally expressed as\\*
\Q@ class A use B,C {...new methods...}@\\*
can be expressed in python as \\*
\Q@ class Aux: ...new methods...@\\*
\Q@ class A(B,C,Aux):  pass@ 
\end{enumerate}

\subsection{Multiple Inheritance in Java}
Since Java 8 default methods are introduced, which enables concrete method
implementation to be defined (via the \textbf{default} keyword) inside
interfaces. Since Java supports implementation of multiple interfaces (instead of extension of a single
class), the introduction of default methods opens the gate for various flavours of multiple
inheritance in Java, using interfaces. 
Former work by Bono.et. al.~\cite{bono14}.
provides details on mimicking traits~\cite{sharly thesis} through interfaces.

There are proposals for extending Java (before Java8) with traits. For example, FeatherTrait
Java (FTJ)~\cite{Liquori08ftj} by Liquori et al. extends the calculus of
Featherweight Java (FJ)~\cite{Igarashi01FJ} with statically-typed traits, adding
trait-based inheritance in Java.
Except for few, mostly syntactic details, their work can be emulated/rephrased with Java8 interface.

\input{sections/Compare.tex}

@@

\section{Future work}

\subsection{Private state/state initialization}
The biggest limitation of our approach is the absence in Java8 of support for private/protected methods in interfaces.That is, in Java8 all members of interfaces are public, including static methods.
Since we use abstract methods to encode the state, our state is always all public; however is impossible for the user to know if a certain method maps directly to a field or if it have a default implementation.
If the use wants a constructor that does not directly maps to the fields, (as for secondary constructors in Scala) he can simply define its own \Q@of@ method and delegate on the generated one, as in
\begin{lstlisting}
@Mixin interface Point{
  int x(); int y();
  static Point of(int val){return Point.of(val,val);}  
  }
\end{lstlisting}
However, the generated \Q@of@ method would also be present and public.
If a future version of Java was to support \emph{static private methods in interfaces} we could extend our code generation to handle also encapsulation.
Currently, is possible to use a public nested class with private static methods inside, but this is ugly and cumbersome. We are considering if our annotation processor can take code with \Q!@Private! annotation and turn it into static private methods of a nested class. In this extension,  also the of method could be made private following the same pattern.

\subsection{Class invariants in ClassLess Java}
Since the objects are created by automatically generate methods,
another limitation of our current approach is that there is no place where the user can dynamically check for class invariants.
In Java often we see code like
\begin{lstlisting}
class Point{ int x; int y;
  Point(int x; int y){this.x=x;this.y=y; assert this.checkInvariant();}
  private boolean checkInvariant(){... x>0,y>0...}
\end{lstlisting}

We are considering an extension of our annotation where 
default methods with the special name \Q@checkInvariant()@ will be called inside of the \Q@of@ methods.
if multiple interfaces are implemented, and more then one offers
\Q@checkInvariant()@,  a composed implementation could be automatically generated, composing by \Q@&&@ the various competing implementations.

\subsection{Clone, toString, equals and hashCode}
Methods originally defined in \Q@Object@, as \Q@clone@ and \Q@toString@ can be supported by our approach, but they need special care. If an interface annotated with \mixin ask an implementation for \Q@clone@, \Q@toString@ \Q@equls@ or \Q@hashCode@ we can easily generate one from the fields.\footnote{In particular, for clone we can do automatic return type refinement as we do for \Q@with-@ and fluent setters. Note how this would solve most of the Java ugliness related to \Q@clone@ methods.}

However, if the user wish to provide its own implementation, since the method is also implemented in \Q@Object@ we would have a conflict, that we have to explicitly resolve inside of \Q@of@, by implementing the method and delegating it to the user implementation, thus

\begin{lstlisting}
@Mixin interface Point{ int x(); int y();
  default Point clone(){ return Point.of(0,0);}//user defined clone
  }
\end{lstlisting} 
Would expand into

\begin{lstlisting}
interface Point{ int x(); int y();
  default Point clone(){ return Point.of(0,0);}//user defined clone
  public static Point of(int _x,int_y){
    return new Point(){...
      public Point clone(){ return Point.super.clone();}
      };  }  }
\end{lstlisting}
